%! TEX root = ../../master.tex
\lecture[]{Mo 09 Mai 2022}{Untitled}

\begin{definition}
  A partially ordered set $I$ is called \vocab{directed},
  if for any $i$, $j\in I$, there is some $k\in I$
  with $i$, $j\leq k$.

  A \vocab{directed limit} is a colimit over a directed set. 
\end{definition}

\begin{example}
  \label{ex:directed-limits-in-r-modules}
  We have a better description for the directed limit
  of $R$-modules over some diagram $(M_i, f_{i\leq j})$
  than
  \autoref{lm:products-and-equalizers-give-limits}:
  \[
    \colim \mathcat{I} = \faktor{\oplus_{i \in I} M_i}{\simeq}
  \]
  where $x\in M_i$ and $y\in M_j$ are equivalent iff
  there exists some $k$ with $i$, $j\leq k$ and
  $f_{i\leq k}(x) = f_{j\leq k}(y) \in M_k$.
  Note that in particular, we can describe every element
  of the colimit by a \emph{sinlge} element $x\in M_i$
  for some $i$, in contrast to a family of elements
  given by the description of
  \autoref{lm:products-and-equalizers-give-limits}.
\end{example}

\begin{definition}
  Let $\mathcat{C}$ be a category and $f\colon X\to Y$
  me a morphism in $\mathcat{C}$.
  We say that $f$ is a \vocab{monomorphism},
  if $f$ is left-cancellable, that is,
  for every object $Z$ and morphisms $g_1$, $g_2\colon Z\to X$,
  if $f\circ g_1 = f \circ g_2$ implies $g_1 = g_2$.
  \[
    \begin{tikzcd}
      Z
      \ar[shift left]{r}{g_1}
      \ar[shift right, swap]{r}{g_2}
      &
      X
      \ar{r}{f}
      &
      Y
    \end{tikzcd}
  .\]
  We also say that $f$ is \vocab{mono} or \vocab{monic}.

  The dual notion is called an \vocab{epimorphism},
  that is, $f$ is \vocab{epic} if $f$ is right-cancellable. 
\end{definition}

\begin{example}+
  The kernel map $ι\colon \ker f \to X$ of a morphism $f\colon X\to Y$
  is always monic.
  If $g_1$, $g_2\colon Z \to  \ker f$ are any two
  maps with $ι \circ g_1 = ι \circ g_2$,
  then in particular $ι \circ  g_1 \circ  f = ι \circ  g_2 \circ f$,
  and by the uniqueness property of the kernel,
  the map $g_1 = g_2$ is the \emph{unique}
  factorization of this map over $\ker f$.

  Dually, the projection map $Y \to  \coker f$
  is always an epimorphism.
\end{example}

\begin{definition}
  \label{def:abelian-category}
  A pre-abelian group is abelian if every monomorphism
  is a kernel and every epimorphism is a cokernel.
\end{definition}


\section{Compactly supported cohomology}

\begin{lemma}
  Let $M$, $N$ and $P$ be three diagrams of $R$-modules,
  indexed over some directed set $\mathcat{I}$.
  Let $u\colon M\to N$ and $v\colon N \to P$ be natural
  transformations between these diagrams.
  Assume that for any $i\in \mathcat{I}$,
  the sequence
  \[
    M_1 \xrightarrow{u_i}
    N_i \xrightarrow{v_i}
    P_i
  \]
  is exact.
  Then also the induced sequence on directed limits
  \[
    \colim_{\mathcat{I}} M
    \to
    \colim_{\mathcat{I}} N
    \to
    \colim_{\mathcat{I}} P
  \]
  is exact.
\end{lemma}
